Determinism, Complexity, and Predictability in Computer Performance

### 2014-04-15 concl.tex
finished draft of conclusion

Forecast strategies that are designed to capture predictive structure
are ineffective when signal complexity outweighs information
redundancy.  This poses a number of serious challenges in practice.
First, without knowing anything about the generating process, it is
difficult to determine how much predictive structure is present in a
given time series.  And even if predictive structure exists, a given
forecast method may not work, simply because it cannot exploit the
structure that is present (e.g., a linear model of a nonlinear
process).  If a forecast model is not producing good results, a
practitioner needs to know why: is the reason that the time series
contains no predictive structure---i.e., that no model will work---or
is the model that s/he is using simply not good enough?

In this paper, we have argued that redundancy~\cite{crutchfield2003}
is a useful definition of the inherent predictability of a time
series.  To operationalize that definition, we use an approximation of
the Kolmogorov-Sinai entropy~\cite{lind95}, estimated using a weighted
version of the permutation entropy of~\cite{bandt2002per}.  This WPE
technique---an ordinal calculation of forward information transfer in
a time series---is ideal for our purposes because it works with
real-valued data and is known to converge to the true entropy value.
Using a variety of forecast models and more than a hundred time-series
data sets from a computer-performance experiment, we have shown that
prediction accuracy is indeed correlated with weighted permutation
entropy: the higher the WPE, in general, the higher the prediction
error.  There are some exceptions: [[Liz to do: summarize \& discuss
    here.]]

@@

An important practical corollary to this empirical correlation of
predictability and WPE is a practical strategy for assessing
appropriateness of forecast methods.  If the forecast produced by a
particular method is poor but the time series contains a significant
amount of predictive structure, one can reasonably conclude that that
method is inadequate to the task and that one should seek another
method.  The nonlinear LMA method, for instance, performs better in
most cases because it is more general.  The \naive ~method, which
simply predicts the mean, works better in noisy signals because it
effects a filtering operation.  The simple random-walk strategy is
better than LMA, ARIMA, or the \naive ~method on the \gcc signal,
which is extremely complex---i.e., extremely low redundancy.
{\color{red} I suggest we don't mention \svdone here since it's a
  pretty detailed thing and conclusions are for high-level summaries.
  It's a good point, though, and I'll leave it in comments in the
  file.}
%  \naive ~wins on \svdone with the exception to the 5 outliers.
The dashed line in Figures~\ref{fig:wpe_vs_mase_best}
and~\ref{fig:wpe_vs_mase_all} is a useful empirical guideline for
knowing when a model is not well-matched to the task at hand: a $MASE$
score that is more than {\color{red}[[what's the equation for that
      line, Ryan?]]} times the WPE of the corresponding signal is an
indication that that time series has more predictive structure than
that forecast model can capture and exploit.

@@

Nonstationarity is a serious challenge in any time-series modeling
problem.  Indeed, one of the first applications of permutation entropy
was regime-shift detection for the purposes of recognizing seizure
onset in brainwave data~\cite{cao2004det}.  Recall that the signal in
Figure~\ref{fig:svd-ts-colored} was especially useful for the study in
this paper because it contained a number of different regimes.  We
segmented this signal visually, but one could imagine using
permutation entropy to do so instead.  Automating regime-shift
detection would be an important step towards a fully adaptive modeling
strategy, where old models are discarded and new ones are rebuilt
whenever shifts are detected.  WPE would be particularly powerful in
this scenario, as its value can suggest what kind of model might work
well in each new regime.

All of the experiments described in this paper involve a particular
physical system.  We chose that system because its behavior spans a
broad spectrum of classical time series patterns.  We believe that our
results will generalize far beyond this particular system; proving
that claim with examples is another important next step.  [[Liz to do:
    insert some examples.  Close with mention of what that could
    enable (viz., adaptive models like ones at end of prev paragraph
    used as the core of controllers).]]

>>>>

Forecast strategies that are designed to capture predictive structure
are ineffective when signal complexity outweighs information
redundancy.  This poses a number of serious challenges in practice.
Without knowing anything about the generating process, it is difficult
to determine how much predictive structure is present in a given time
series.  And even if predictive structure exists, a given forecast
method may not work, simply because it cannot exploit the structure
that is present (e.g., a linear model of a nonlinear process).  If a
forecast model is not producing good results, a practitioner needs to
know why: is the reason that the time series contains no predictive
structure---i.e., that no model will work---or is the model that s/he
is using simply not good enough?

In this paper, we have argued that redundancy~\cite{crutchfield2003}
is a useful definition of the inherent predictability of a time
series.  To operationalize that definition, we use an approximation of
the Kolmogorov-Sinai entropy~\cite{lind95}, estimated using a weighted
version of the permutation entropy of~\cite{bandt2002per}.  This WPE
technique---an ordinal calculation of forward information transfer in
a time series---is ideal for our purposes because it works with
real-valued data and is known to converge to the true entropy value.
Using a variety of forecast models and more than a hundred time-series
data sets from a computer-performance experiment, we have shown that
prediction accuracy is indeed correlated with weighted permutation
entropy: the higher the WPE, in general, the higher the prediction
error.  There are some exceptions: \svdone is easier to predict than
its WPE value suggests using any of the three forecast strategies
studied here.  Conversely, forecast errors for the \col and \svdfive
signals are higher than the corresponding WPEs suggest---except for
the nonlinear LMA predictor.

@@

An important practical corollary to this empirical correlation of
predictability and WPE is a practical strategy for assessing
appropriateness of forecast methods.  If the forecast produced by a
particular method is poor but the time series contains a significant
amount of predictive structure, one can reasonably conclude that that
method is inadequate to the task and that one should seek another
method.  The nonlinear LMA method, for instance, performs better in
most cases because it is more general.  (This is particularly apparent
in the \col and \svdfive examples.)
% , where the other two methods do not perform well.)  
The \naive ~method, which simply predicts the mean, can work very well
on noisy signals because it effects a filtering operation.  The simple
random-walk strategy outperforms LMA, ARIMA, and the \naive ~method on
the \gcc signal, which is extremely complex---i.e., extremely low
redundancy.
%  \naive ~wins on \svdone with the exception to the 5 outliers.
The dashed line in Figures~\ref{fig:wpe_vs_mase_best}
and~\ref{fig:wpe_vs_mase_all} operationalizes the discussion in this
paragraph.  It is a useful empirical guideline for knowing when a
model is not well-matched to the task at hand: a MASE score that is
more than {\color{red}[[what's the equation for that line, Ryan?]]}
times the WPE of the corresponding signal is an indication that that
time series has more predictive structure than that forecast model can
capture and exploit.

@@

Nonstationarity is a serious challenge in any time-series modeling
problem.  Indeed, one of the first applications of permutation entropy
was to recognize the regime shift in brainwave data that occurs when
someone has a seizure~\cite{cao2004det}.  Recall that the signal in
Figure~\ref{fig:svd-ts-colored} was especially useful for the study in
this paper because it contained a number of different regimes.  We
segmented this signal visually, but one could imagine using
permutation entropy to do so instead.  Automating regime-shift
detection would be an important step towards a fully adaptive modeling
strategy, where old models are discarded and new ones are rebuilt
whenever the time series enters a new regime.  WPE would be
particularly powerful in this scenario, as its value can not only help
with regime-shift detection, but also suggest what kind of model might
work well in each new regime.

All of the experiments described in this paper involve a particular
physical system.  We chose that system because its behavior spans a
broad spectrum of classical time series patterns.  Since these
patterns---dynamical chaos, periodicities, noise, etc.--- are generic,
we believe that our results will generalize beyond this particular
system.  Testing that claim with different examples is another
important next step.  Of particular interest would be the class of
so-called \emph{hybrid systems}: [[insert def]].  A lathe, for
instance, that has an intermittent instability.  Traffic at an
internet router, which has characteristic patterns during normal
operation that shift radically during an attack.  Modeling and
predicting the behavior of these kinds of systems effectively is quite
difficult; doing so adaptively and automatically---in the manner that
is alluded to at the end of the previous paragraph---would be an even
more important challenge.

### 2014-04-15 model.tex
typo in fig labels
finished draft of conclusion

      \begin{subfigure}{0.6\columnwidth}
    \includegraphics[width=\columnwidth]{figs/colLMAForecast.png}
    \caption{\col \\ LMA}
    \label{fig:colLMA}
  \end{subfigure}
      \begin{subfigure}{0.6\columnwidth}
    \includegraphics[width=\columnwidth]{figs/gccLMAForecast.png}
    \caption{\gcc \\ LMA}
    \label{fig:gccLMA}
  \end{subfigure}  
        \begin{subfigure}{0.6\columnwidth}
    \includegraphics[width=\columnwidth]{figs/svdfiveLMAForecast.png}
    \caption{\svdfive \\ LMA}
    \label{fig:gccLMA}
  \end{subfigure}
  %\begin{subfigure}{0.5\textwidth}
  %  \includegraphics[width=1.0\textwidth]{figs/LMA_vs_ARIMA}
   \caption{Predicted ($p_j$) versus true values ($c_j$) for
     forecasts of \col, \gcc, and \svdfive.  On this type of plot, a
     perfect prediction lies exactly on the diagonal.
%
%
}
\label{fig:forecast-example}
  %\end{subfigure}
\end{figure*} 
In these images, the vertical axis is the prediction $p_j$ and the
horizontal axis is the true continuation $c_j$.  On such a plot, a
perfect prediction would lie on the diagonal.  LMA, for instance,
generates a very accurate prediction of the \col data, while ARIMA
does not.  Horizontal lines result when a constant predictor (e.g.,
\naive) is used on a non-constant signal.  Point clouds reflect the
structure of the distribution of the errors---roughly normally
distributed, for example, in the case of LMA on \gcc.  The shapes of
Figures~\ref{fig:colRW} and~\ref{fig:colARIMA} are reminiscent of the
projected embedding in Figure~\ref{fig:embedding}(a).  For a
random-walk predictor, a $p_j$ vs. $c_j$ plot is equivalent to a
two-dimensional embedding with $\tau=1$.  For ARIMA, the
correspondence is not quite as simple, since the $p_j$ values are
linear combinations of a number of past values of the $c_j$, but the
effect is largely the same\footnote{The temporal ordering of the
  points on the ARIMA $p_j$ vs. $c_j$ plot does not match that of a
  projected embedding of the time series, however.}.

As a numerical measure of prediction accurary, we calculate the mean
absolute scaled error ($MASE$) between the true and predicted signals:
% 
$$MASE = \sum_{j=n+1}^{k+n+1}\frac{|p_j-c_j|
}{\frac{k}{n-1}\sum^n_{i=2}|x_{i}-x_{i-1}|}$$
% 
This error metric was introduced in \cite{MASE} as a ``generally
applicable measurement of forecast accuracy without the problems seen
in the other measurements."  $MASE$ is a normalized measure; the
scaling term in the denominator
% 
% of the MASE value
% $$\frac{1}{n-1}\sum^n_{i=2}|X_{i,obs}-X_{i-1,obs}|$$ 
% 
is the average in-sample forecast error for a random-walk prediction
over the initial training signal $\{x_i\}^n_{i=1}$.  That is, $MASE<1$ means that the
prediction error in question was, on the average, smaller than the
error of a random-walk forecast on the same data.  Analogously,
$MASE>1$ means that the corresponding prediction method did
\emph{worse}, on average, than the random-walk method.  We chose this
error metric because it allows for fair comparison across varying
methods, prediction horizons, and signal scales.

$MASE$ scores for all 360 experiments are tabulated in the three
middle columns in Table~\ref{tab:error}.
\begin{table*}
  \begin{center}
  \begin{tabular}{|c|c|c|c|c|}
  \hline Signal & LMA  & ARIMA & na\"{i}ve  & Weighted Permutation \\ 
  &  $MASE$ & $MASE$  & $MASE$ &  Entropy \\ \hline
 \col           & $ 0.050 \pm0.002  $ & $0.599  \pm 0.211 $ & $0.571\pm0.002$&  $0.513 \pm 0.003$ \\

@@

\svdfive     & $ 0.718\pm 0.048 $ & $2.370  \pm 0.051 $ & $20.870 \pm 0.192$&  $0.678 \pm 0.007$ \\
 \svdsix     & $ 0.739\pm 0.068 $ & $ 1.438\pm 0.061$ & $2.197\pm0.083$&  $0.748 \pm 0.011$ \\
  \hline
  \end{tabular}
  \end{center}
 \label{default}
 \caption{$MASE$ scores and weighted permutation entropies for all eight
   processes studied in this paper}
 \label{tab:error}
  \end{table*}%
Comparing these with the geometry of the plots in
Figure~\ref{fig:forecast-example}, one can see some obvious
correspondences.  The average LMA $MASE$ score for the \col signals
was $0.050$, for instance, while ARIMA scored much worse ($0.599$).
That is, LMA performed roughly 20 times better on \col signals than a
random-walk predictor, while ARIMA only outperformed random walk by a
factor of 1.7.  This is in accordance with the visual appearance of
the corresponding images in Figure~\ref{fig:forecast-example}.  While
the correspondence between $MASE$ score and the visual appearance of
these kinds of plots is clear cut in this case, that is not always so.
The plots of LMA predictions of \col and \svdfive both lie near the
diagonal, for instance, but the corresponding $MASE$ scores were very
different: $0.050$ for \col and $0.718$ for \svdfive (resp., 20 and
1.4 times better than random-walk forecasts of the same signals).
This is a result of the normalization in the $MASE$ score calculation.
Recall from Section~\ref{sec:simple} that the random-walk method
performs especially poorly on signals that oscillate rapidly.  \col
fits this description, so the random-walk error on this signal---the
denominator of the three $MASE$ scores in the first row of the
Table---is pathologically high, which skews those scores down.
Random-walk prediction is very effective for the \svdfive signal, on
the other hand, so the denominator is small and the $MASE$ scores are
skewed upwards.  Normalization is, as is often the case, a double-edged
sword.  Here, it facilitates comparisons across signals of different
types and lengths, but its particular pathologies must be taken into
account, as discussed at more length in Section~\ref{sec:results}.

% Table \ref{tab:error} provides the $MASE$ distributions
% {\color{red}[[Joshua: Ryan, Is this the right word? we give mean $\pm$
%       std. dev but some have very skewed right tails]]} for each of
% the eight signals and three prediction strategies, these are averaged
% over 15 runs of each signal-method combination. For later comparison,
% Table \ref{tab:error} also has the complexity measure we introduce in
% Section \ref{sec:meaComplex}.

>>>>

      \begin{subfigure}{0.6\columnwidth}
    \includegraphics[width=\columnwidth]{figs/colLMAForecast.png}
    \caption{\col \\ LMA}
    \label{fig:colLMA}
  \end{subfigure}
      \begin{subfigure}{0.6\columnwidth}
    \includegraphics[width=\columnwidth]{figs/gccLMAForecast.png}
    \caption{\gcc \\ LMA}
    \label{fig:gccLMA}
  \end{subfigure}  
        \begin{subfigure}{0.6\columnwidth}
    \includegraphics[width=\columnwidth]{figs/svdfiveLMAForecast.png}
    \caption{\svdfive \\ LMA}
    \label{fig:svd5LMA}
  \end{subfigure}
  %\begin{subfigure}{0.5\textwidth}
  %  \includegraphics[width=1.0\textwidth]{figs/LMA_vs_ARIMA}
   \caption{Predicted ($p_j$) versus true values ($c_j$) for
     forecasts of \col, \gcc, and \svdfive.  On this type of plot, a
     perfect prediction lies exactly on the diagonal.
%
%
}
\label{fig:forecast-example}
  %\end{subfigure}
\end{figure*} 
In these images, the vertical axis is the prediction $p_j$ and the
horizontal axis is the true continuation $c_j$.  On such a plot, a
perfect prediction would lie on the diagonal.  LMA, for instance,
generates a very accurate prediction of the \col data, while ARIMA
does not.  Horizontal lines result when a constant predictor (e.g.,
\naive) is used on a non-constant signal.  Point clouds reflect the
structure of the distribution of the errors---roughly normally
distributed, for example, in the case of LMA on \gcc.  The shapes of
Figures~\ref{fig:colRW} and~\ref{fig:colARIMA} are reminiscent of the
projected embedding in Figure~\ref{fig:embedding}(a).  For a
random-walk predictor, a $p_j$ vs. $c_j$ plot is equivalent to a
two-dimensional embedding with $\tau=1$.  For ARIMA, the
correspondence is not quite as simple, since the $p_j$ values are
linear combinations of a number of past values of the $c_j$, but the
effect is largely the same\footnote{The temporal ordering of the
  points on the ARIMA $p_j$ vs. $c_j$ plot does not match that of a
  projected embedding of the time series, however.}.

As a numerical measure of prediction accurary, we calculate the mean
absolute scaled error (MASE) between the true and predicted signals:
% 
$$MASE = \sum_{j=n+1}^{k+n+1}\frac{|p_j-c_j|
}{\frac{k}{n-1}\sum^n_{i=2}|x_{i}-x_{i-1}|}$$
% 
This error metric was introduced in \cite{MASE} as a ``generally
applicable measurement of forecast accuracy without the problems seen
in the other measurements."  MASE is a normalized measure; the
scaling term in the denominator
% 
% of the MASE value
% $$\frac{1}{n-1}\sum^n_{i=2}|X_{i,obs}-X_{i-1,obs}|$$ 
% 
is the average in-sample forecast error for a random-walk prediction
over the initial training signal $\{x_i\}^n_{i=1}$.  That is, $MASE<1$ means that the
prediction error in question was, on the average, smaller than the
error of a random-walk forecast on the same data.  Analogously,
$MASE>1$ means that the corresponding prediction method did
\emph{worse}, on average, than the random-walk method.  We chose this
error metric because it allows for fair comparison across varying
methods, prediction horizons, and signal scales.

MASE scores for all 360 experiments are tabulated in the three
middle columns in Table~\ref{tab:error}.
\begin{table*}
  \begin{center}
  \begin{tabular}{|c|c|c|c|c|}
  \hline Signal & LMA  & ARIMA & na\"{i}ve  & Weighted Permutation \\ 
  &  MASE & MASE  & MASE &  Entropy \\ \hline
 \col           & $ 0.050 \pm0.002  $ & $0.599  \pm 0.211 $ & $0.571\pm0.002$&  $0.513 \pm 0.003$ \\

@@

\svdfive     & $ 0.718\pm 0.048 $ & $2.370  \pm 0.051 $ & $20.870 \pm 0.192$&  $0.678 \pm 0.007$ \\
 \svdsix     & $ 0.739\pm 0.068 $ & $ 1.438\pm 0.061$ & $2.197\pm0.083$&  $0.748 \pm 0.011$ \\
  \hline
  \end{tabular}
  \end{center}
 \label{default}
 \caption{MASE scores and weighted permutation entropies for all eight
   processes studied in this paper}
 \label{tab:error}
  \end{table*}%
Comparing these with the geometry of the plots in
Figure~\ref{fig:forecast-example}, one can see some obvious
correspondences.  The average LMA MASE score for the \col signals
was $0.050$, for instance, while ARIMA scored much worse ($0.599$).
That is, LMA performed roughly 20 times better on \col signals than a
random-walk predictor, while ARIMA only outperformed random walk by a
factor of 1.7.  This is in accordance with the visual appearance of
the corresponding images in Figure~\ref{fig:forecast-example}.  While
the correspondence between MASE score and the visual appearance of
these kinds of plots is clear cut in this case, that is not always so.
The plots of LMA predictions of \col and \svdfive both lie near the
diagonal, for instance, but the corresponding MASE scores were very
different: $0.050$ for \col and $0.718$ for \svdfive (resp., 20 and
1.4 times better than random-walk forecasts of the same signals).
This is a result of the normalization in the MASE score calculation.
Recall from Section~\ref{sec:simple} that the random-walk method
performs especially poorly on signals that oscillate rapidly.  \col
fits this description, so the random-walk error on this signal---the
denominator of the three MASE scores in the first row of the
Table---is pathologically high, which skews those scores down.
Random-walk prediction is very effective for the \svdfive signal, on
the other hand, so the denominator is small and the MASE scores are
skewed upwards.  Normalization is, as is often the case, a double-edged
sword.  Here, it facilitates comparisons across signals of different
types and lengths, but its particular pathologies must be taken into
account, as discussed at more length in Section~\ref{sec:results}.

% Table \ref{tab:error} provides the MASE distributions
% {\color{red}[[Joshua: Ryan, Is this the right word? we give mean $\pm$
%       std. dev but some have very skewed right tails]]} for each of
% the eight signals and three prediction strategies, these are averaged
% over 15 runs of each signal-method combination. For later comparison,
% Table \ref{tab:error} also has the complexity measure we introduce in
% Section \ref{sec:meaComplex}.
